---
output: 
  pdf_document
title: "Dietary trait QC for UKB gene-diet interaction analysis"
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F, message=F,warning=F, 
                      fig.path="../output/figures/2b_diet_qc/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "cowplot", 
    "pheatmap", "RColorBrewer"), 
  library, character.only=T))
```

```{r load-data}
diet_input <- read_tsv("../data/processed/final_baseline_ffq_data.txt")

kw_pca <- readRDS("../data/processed/ffq_pca_fit.rds")
kw_pca_jbcFields <- readRDS("../data/processed/ffq_pca_fit_jbcFields.rds")

kwPCs <- read_tsv("../data/processed/ffq_PCs.txt")
jcPCs <- read_tsv("../data/processed/JBC_BOLT_UKB_diet_genoQCEUR450K_170FFQphenotypes_agesexadj_INV") %>%
  rename_all(~gsub("median_", "", .)) %>%
  select(id=IID, everything(), -FID)

diet_BL <- read_tsv("../data/processed/diet_baseline.txt")
```

# Input dietary variables

* These correlations are based on the final PCA input dataset (i.e. after INT and median imputation)

```{r ffq-trait-correlations, fig.asp=1}
diet_input_corrs <- cor(select(diet_input, -id, -sex, -age_months),
                        use="pairwise.complete.obs")
diet_input_corrs[is.na(diet_input_corrs)] <- 0  # Deals with milk QT vs. binary trait issue
rownames(diet_input_corrs) <- strtrim(rownames(diet_input_corrs), 15)
colnames(diet_input_corrs) <- strtrim(colnames(diet_input_corrs), 25)
pheatmap(
  diet_input_corrs,
  breaks=seq(-max(abs(diet_input_corrs)), max(abs(diet_input_corrs)), 
             length.out=100),
  treeheight_row=0, treeheight_col=0,
  angle_col=90,
  main="Pearson correlations between dietary traits"
)
```

\newpage

# Dietary PCs

```{r scree}
qplot(x=1:30,
      y=kw_pca$sdev ** 2 / sum(kw_pca$sdev ** 2) * 100) +
  geom_line() + 
  scale_x_continuous(breaks=c(1, seq(5, 30, 5))) +
  labs(x="PC", y="% variance explained",
       title="Scree plot for dPCs")
```


```{r loadings}
kw_loadings <- kw_pca$rotation
rownames(kw_loadings) <- strtrim(rownames(kw_loadings), 15)
pheatmap(
  kw_loadings,
  breaks=seq(-max(abs(kw_loadings)), max(abs(kw_loadings)), length.out=100),
  cluster_rows=F, cluster_cols=F,
  angle_col=45,
  main="Dietary principal component loadings"
)
```

\newpage

# Comparison of KW and JC PCs

* Differences may have arisen due to any of the following changes:
    - Sample inclusion:
        + Only unrelated Europeans
        + Inclusion of disease & pregnancy
    - Baseline FFQ responses only
    - Alcohol imputation -- JC imputed values in {-1, -3} as the median of the full set of non-missing values, versus KW as the median of the values not in {-1, -3}
    - No residualizing on age + sex
    - Trimming to a much smaller set of binary variables to reduce heavy collinearity

```{r compare-PCs}
combined <- jcPCs %>%
  select(id, paste0("ffq_PC", 1:5)) %>%
  setNames(c("id", paste0("ffq_PC", 1:5, ".JBC"))) %>%
  inner_join(kwPCs, by="id") %>%
  setNames(gsub("ffq_", "", names(.)))

pc_corrs <- cor(select(combined, -id))
pc_corrs <- pc_corrs[grepl("JBC", rownames(pc_corrs)),
                     !grepl("JBC", colnames(pc_corrs))]

pheatmap(
  pc_corrs,
  breaks=seq(-max(abs(pc_corrs)), max(abs(pc_corrs)), length.out=100),
  cluster_rows=F, cluster_cols=F,
  angle_col=45,
  main="Correlations with top 5 JBC dietary PCs"
)

# kw_pca_fit <- readRDS("../data/processed/ffq_pca_fit_jbcFields.rds")
# load("/humgen/diabetes2/users/jcole/UKBB/diet/ffq_median.pca_11212018update.Rdata")
# jc_pca_fit <- ffq_median.pca
# 
# plot(x=kw_pca_fit$rot[, 25], y=jc_pca_fit$rot[, 25],
#      xlab="KW loadings", ylab="JC loadings",
#      main="Comparison of PC25 loadings")
# text(x=kw_pca_fit$rot[, 25], y=jc_pca_fit$rot[, 25],
#      labels=rownames(kw_pca_fit$rot),
#      cex=0.3)
```

\newpage

# Rough QC of KW preprocessing based on missingness

* Uses the "raw" dietary dataset (i.e. missingness calculated prior to imputation)

```{r check-missingness}
pca_fields <- c(
  "id",
  "selfreportsex.31",
  "age_months",
  "cookedveg_TBSperday.1289_QT",
  "rawveg_TBSperday.1299_QT",
  "freshfruit_piecesperday.1309_QT",
  "driedfruit_piecesperday.1319_QT",
  "bread_slicesperweek.1438_QT",
  "cereal_bowlsperweek.1458_QT",
  "tea_cupsperday.1488_QT",
  "coffee_cupsperday.1498_QT",
  "water_glassesperday.1528_QT",
  "anyalcohol_glassespermonth.derived_QT",
  "oilyfish_overallfreq.1329_QT",
  "nonoilyfish_overallfreq.1339_QT",
  "processmeat_overallfreq.1349_QT",
  "poultry_overallfreq.1359_QT",
  "beef_overallfreq.1369_QT",
  "lambmutton_overallfreq.1379_QT",
  "pork_overallfreq.1389_QT",
  "cheese_overallfreq.1408_QT",
  "doyouaddsalt.1478_QT",
  "hotdrinktemp.1518_QT",
  "milk_typeused.1418_QT14",
  "nevereatcategories.6144_bin2",
  "nevereatcategories.6144_bin4",
  "nevereatcategories.6144_bin6",
  "nevereatcategories.6144_bin8",
  "milk_typeused.1418_bin",
  "spread_typeused.1428.0.nonbutterspread_typeused.2654.combined_bin7",
  "bread_typeused.1448_bin4",
  "cereal_typeused.1468_bin5",
  "coffee_typeused.1508_bin3"
)

missing_tbl <- lapply(pca_fields, function(f) round(sum(!is.na(diet_BL[[f]])) / nrow(diet_BL), 2))
names(missing_tbl) <- pca_fields
# saveRDS(missing_tbl, "../data/processed/sanity_check_fields.rds")
# missing_tbl <- readRDS("../data/processed/sanity_check_fields.rds")
kw_missingness <- tibble(trait=names(missing_tbl), frac=unlist(missing_tbl))

jc_missing_tbl <- read_tsv("../data/raw/UKB_diet_SampleSizeByTrait.txt")
jc_missingness <- tibble(trait=jc_missing_tbl$trait, 
                         frac=round(jc_missing_tbl$N / 449210, 2))
jc_missingness2 <- mutate(jc_missingness, trait=gsub("\\.average", "", trait))

compare_df <- full_join(kw_missingness, jc_missingness2, 
                        by="trait", suffix=c("_kw", "_jc")) %>%
  filter(!is.na(frac_kw),
         grepl("_QT|_bin", trait))

kable(compare_df, longtable=T,
      caption="Fractions non-missing in KW vs. JC diet traits")
```

\newpage

# Histograms for each trait (prior to median imputation & INT)

```{r histograms}
plt_df <- mutate_at(diet_BL, vars(contains("_bin")), 
                    function(x) as.factor(ifelse(is.na(x), NA, round(x, 2))))
histograms <- lapply(names(plt_df)[-1], function(nm) {
  pct_nonmissing <- round(sum(!is.na(plt_df[[nm]])) / nrow(plt_df) * 100, 1)
  string_nonmissing <- paste0(nm, " (", pct_nonmissing, "% non-missing)")
  if (grepl("_bin|sex", nm)) {
    ggplot(plt_df, aes_string(x=nm)) +
      geom_bar(stat="count") +
      labs(x=string_nonmissing)
  } else {
    ggplot(plt_df, aes_string(x=nm)) +
      geom_histogram() +
      labs(x=string_nonmissing)
  }
})

silent <- lapply(histograms, print)
```
