---
output: 
  pdf_document:
    latex_engine: pdflatex
title: "Phenotype QC for UKB gene-diet interaction analysis"
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F, message=F,warning=F, fig.path="../output/figures/")
suppressMessages(silent <- lapply(
  c("knitr", "kableExtra", "tidyverse", "cowplot", "jtools"), 
  library, character.only=T))
```

```{r auxiliary}
INT <- function(x) qnorm((rank(x, na.last="keep") - 0.5) / sum(!is.na(x)))
```

```{r read-phenos}
raw_phenos <- read_csv("../data/processed/ukbb_diet_gwis_phenos_raw.csv")
phenos <- read_csv("../data/processed/ukbb_diet_gwis_phenos.csv")
```

All data shown below are from unrelated, European-ancestry individuals in UK Biobank who have not revoked consent for analysis (N ~ 350k). HbA1c distributions then removed individuals with confirmed diabetes (T1D or T2D based on Florez project "prob_poss_..." definitions) and outliers (outside of 3 IQRs from the 25th/75th percentiles) unless otherwise noted.

```{r distributions}
hba1c_hist <- ggplot(phenos, aes(x=hba1c_raw)) + 
  geom_histogram() +
  labs(title="HbA1c raw")
hba1c_INT_hist <- ggplot(phenos, aes(x=hba1c_INT)) +
  geom_histogram() +
  labs(title="HbA1c (INT)")
plot_grid(hba1c_hist, hba1c_INT_hist)

top4_ffq_pc_plts <- lapply(1:4, function(i) {
  ggplot(phenos, aes_string(x=paste0("ffq_PC", i))) +
    geom_histogram() +
    labs(title=paste0("FFQ-PC", i), x="")
})
do.call(plot_grid, top4_ffq_pc_plts)

ggplot(raw_phenos, aes(x=sex)) +
  geom_bar()

ggplot(phenos, aes(x=age)) +
  geom_histogram()

ggplot(raw_phenos, aes(x=fct_rev(assessment_centre))) +
  geom_bar() +
  coord_flip() +
  labs(x="Assessment Centre")

ggplot(raw_phenos, aes(x=fct_rev(birthplace))) +
  geom_bar() +
  coord_flip() +
  labs(x="Birthplace")
```

## Are potential confounders (birthplace, assessment center) associated with HbA1c?

```{r hba1c-associations}
hba1c_ac_lm <- lm(hba1c_raw ~ assessment_centre, 
                  data=filter(raw_phenos, !(t1d | t2d)))
plot_summs(hba1c_ac_lm)

hba1c_bp_lm <- lm(hba1c_raw ~ birthplace, 
                  data=filter(raw_phenos, !(t1d | t2d)))
plot_summs(hba1c_bp_lm)

pc_string <- paste0("PC", 1:10, collapse=" + ")
hba1c_pcs_lm <- lm(as.formula(paste0("hba1c_raw ~ ", pc_string)), data=raw_phenos)
plot_summs(hba1c_pcs_lm)

hba1c_basic_lm <- lm(hba1c_raw ~ age + sex + age_squared, data=raw_phenos)
plot_summs(hba1c_basic_lm)

all_covar_string <- paste0(pc_string, " + assessment_centre + birthplace")
hba1c_all_lm <- lm(as.formula(paste0("hba1c_raw ~ ", all_covar_string)), data=raw_phenos)
plot_summs(hba1c_all_lm)
```

## Histograms for the full set of 39 FFQ-PCs

```{r all-ffq-pc-histograms}
all_ffq_PCs <- grep("ffq_PC", names(phenos), value=T)
ffq_pc_plts <- lapply(all_ffq_PCs, function(pc) {
  ggplot(phenos, aes_string(x=pc)) +
    geom_histogram() +
    labs(title=gsub("ffq_PC_", "FFQ-PC ", pc), x="")
})
for (plt in ffq_pc_plts) print(plt)
```

```{r additional-variables, eval=F}
acbpAdj_lm <- lm(hba1c_raw ~ assessment_centre + birthplace, data=raw_phenos,
                 na.action=na.exclude)
phenos_expanded <- phenos %>%
  mutate(ffq_median_PC1_INT=INT(ffq_PC1),
         hba1c_acbpResid_INT=INT(resid(acbpAdj_lm)))
write_csv(phenos_expanded, "../data/processed/ukbb_diet_gwis_phenos_expanded.csv")
```